Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2023-06-20 manuscript.Rmd
Added new figure

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink [@McDougal2008]. There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; @Wyatt1981). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus [@Angee2021; @Mathot2018; @McDougal2008; @Wang2015]. To account for the consensual response, these brain regions must combine information from the left and right eyes [@Doesschate1967], yet the computation that achieves this is achieved is unclear.

@@

The stimuli were two discs of achromatic flickering light with a diameter of 3.74 degrees, presented on a black background. The same stimuli were used for all three experiments. Four dark red lines were added around both discs to help with their perceptual fusion, giving the appearance of a single binocular disc (see insert in Figure \@ref(fig:pupildata)b for an example of the fused stimulus). The discs were viewed through a four-mirror stereoscope, which used front silvered mirrors to avoid internal reflections, and meant that participants saw a single fused disc. The use of a stereoscope allowed us to modulate the stimuli in three different ocular configurations: monocular, binocular, and dichoptic. Note that during monocular presentation of flicker, the unstimulated eye still saw the static (non-flickering) disc of mean luminance.

@@

We included six distinct ocular conditions, each at five temporal contrast levels (combined factorially) relative to the mean luminance: 6, 12, 24, 48 and 96\%. Contrast was defined as temporal Michelson contrast; the difference between maximum and minimum luminances, scaled by the mean and expressed as a percentage. In the first three conditions, the discs flickered at 2 Hz, in either a monocular, binocular, or dichoptic arrangement. In the dichoptic condition the non-target eye saw a fixed temporal contrast of 48\%. In the remaining three conditions (termed the cross-frequency conditions) one eye's disc flickered at 1.6Hz, and the other eye's disc flickered at 2Hz. We also tested monocular (one eye sees 1.6Hz flicker, the other sees mean luminance), binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 2Hz, mask contrast of 48\% at 1.6Hz in the other eye) arrangements. The rationale for flickering both eyes at 2Hz is that we can then measure summation behaviour between the eyes in the pupil and EEG response at 2Hz. The rationale for flickering the eyes at different frequencies is that this permits measurement of suppression between the eyes (i.e. the reduction in the 2Hz response when a 1.6Hz mask component is added to the other eye). We counterbalanced presentation of the target stimulus across the left and right eyes.

@@

```{r createfig1, include=FALSE, results='hide'}

@@

```{r pupildata, fig.cap="Summary of pupillometry results for N=30 participants. Panel (a) shows a group average waveform for binocular presentation (low pass filtered at 5Hz), with the driving signal plotted at the foot. Negative values indicate constriction relative to baseline, and positive values indicate dilation. Panel (b) shows the average Fourier spectrum (absolute amplitude values), with an inset image illustrating the stimulus appearance (upper right). Panels (c,d) show contrast response functions for pupil diameter at 2Hz for different conditions. Panel (e) shows contrast response functions at 1.6Hz for three conditions. Shaded regions and error bars indicate bootstrapped standard errors.", fig.align="center", echo=FALSE}

@@

```{r createfig2, include=FALSE, results='hide'}

@@

```{r createfig3, include=FALSE, results='hide'}

@@

```{r createfig4, include=FALSE, results='hide'}

@@

The results from the matching experiment are shown in Figure \@ref(fig:matchingdata). Each data point indicates the contrast levels required in each eye that were perceptually equivalent to the binocular 24\% (red circles) and 48\% (blue circles) matching contrasts. At both matching contrasts, we see a very substantial increase in the physical contrast required for a monocular target (data points along the x- and y-axes), compared to a binocular target (points along the diagonal of x=y). For example with a 48\% match, the monocular targets required contrasts close to 100\%, whereas binocular targets required a contrast of around 50\%. The data points between these extremes also fall close to the predictions of a linear summation model (diagonal dotted lines), and are inconsistent with a winner-takes-all (or MAX) model (dashed lines). Overall, these matching results are consistent with the approximately linear summation effects observed in the EEG data of Experiment 1 (Figure \@ref(fig:EEGdata)c,f).

>>>>

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink [@McDougal2008]. There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; @Wyatt1981). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus [@Angee2021; @Mathot2018; @McDougal2008; @Wang2015]. To account for the consensual response, these brain regions must combine information from the left and right eyes [@Doesschate1967], yet the computation that achieves this is achieved is unclear. The pupil response is temporally low-pass [@Spitschan2014], most likely due to the mechanical limitations of the iris sphincter and dilator muscles.

@@

```{r createmethodsfig, include=FALSE, results='hide'}

if (processdata>0){

pdf("Figures/methodsfig.pdf", bg="transparent", height = 8, width = 8)

contlist <- c(0.06,0.12,0.24,0.48,0.96)
amp <- 0.5
t <- seq(0,1,length=100)
sinetarget <- sin(2 * t * 2*pi)*amp
sinemask <- sin(1.6 * t * 2*pi)*amp

colpal <- c('#FE5000','#8783CF','#228B22','#808080','#6d008b','#8B8000')
symbollist <- c(21:23,21,24,25)
ramp <- colorRamp(c("white", "black"))  # create a ramp from one colour to another
greypal <- rgb(ramp(seq(0, 1, length = 5)), max = 255)   # index the ramp at ten points

plotlims <- c(0.5,4.1,0.5,7)  # define the x and y limits of the plot (minx,maxx,miny,maxy)

par(mar=c(0.1,0.1,0.1,0.1))
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])   # create an empty axis of the correct dimensions

condlabels <- c('Monocular','Binocular','Dichoptic','Monocular cross','Binocular cross','Dichoptic cross')
text(1.2,6:1,condlabels,pos=2,cex=1.2)

text(c(2,3.5),7,c('Left eye','Right eye'),cex=1.6)

points(rep(1.3,6),6:1, pch=symbollist, bg=colpal, cex=1.6, lwd=3)

for (cont in 5:1){

  polygon(c(0,t,t[100])+1.5, c(0,sinetarget*contlist[cont],0) + 6, col=greypal[cont], border=NA)
  polygon(c(0,t,t[100])+1.5, c(0,sinetarget*contlist[cont],0) + 5, col=greypal[cont], border=NA)
  polygon(c(0,t,t[100])+1.5, c(0,sinetarget*contlist[cont],0) + 4, col=greypal[cont], border=NA)
  polygon(c(0,t,t[100])+1.5, c(0,sinetarget*contlist[cont],0) + 2, col=greypal[cont], border=NA)
  polygon(c(0,t,t[100])+1.5, c(0,sinetarget*contlist[cont],0) + 1, col=greypal[cont], border=NA)

  polygon(c(0,t,t[100])+3, c(0,sinetarget*contlist[cont],0) + 5, col=greypal[cont], border=NA)
  polygon(c(0,t,t[100])+3, c(0,sinemask*contlist[cont],0) + 3, col=greypal[cont], border=NA)
  polygon(c(0,t,t[100])+3, c(0,sinemask*contlist[cont],0) + 2, col=greypal[cont], border=NA)

for (cond in 1:6){lines(c(1.5,2.5),c(cond,cond),lty=2)}
for (cond in 1:6){lines(c(3,4),c(cond,cond),lty=2)}

lines(t+3, sinetarget*0.48 + 4, lwd=2)
lines(t+3, sinemask*0.48 + 1, lwd=2)

lines(t+3, sinetarget*0 + 6, lwd=2)
lines(t+1.5, sinetarget*0 + 3, lwd=2)

text(2.6,c(1,2,4,5,6),'2Hz',srt=270)
text(4.1,c(4,5),'2Hz',srt=270)
text(4.1,c(1,2,3),'1.6Hz',srt=270)

text(2.6,3,'Mean',srt=270)
text(4.1,6,'Mean',srt=270)

imwidth <- 0.5
xstart <- 0.4
ystart <- 6.2
rasterImage(stim,xstart,ystart,xstart+imwidth,ystart+imwidth*1.9)

The stimuli were two discs of achromatic flickering light with a diameter of 3.74 degrees, presented on a black background. The same stimuli were used for all three experiments. Four dark red lines were added around both discs to help with their perceptual fusion, giving the appearance of a single binocular disc (see upper left insert in Figure \@ref(fig:methodsfig) for an example of the fused stimulus). The discs were viewed through a four-mirror stereoscope, which used front silvered mirrors to avoid internal reflections, and meant that participants saw a single fused disc. The use of a stereoscope allowed us to modulate the stimuli in three different ocular configurations: monocular, binocular, and dichoptic. Note that during monocular presentation of flicker, the unstimulated eye still saw the static (non-flickering) disc of mean luminance.

@@

We included six distinct ocular conditions, each at five temporal contrast levels (combined factorially) relative to the mean luminance: 6, 12, 24, 48 and 96\%. Contrast was defined as temporal Michelson contrast; the difference between maximum and minimum luminances, scaled by the mean and expressed as a percentage. In the first three conditions, the discs flickered at 2 Hz, in either a monocular, binocular, or dichoptic arrangement (see upper rows of Figure \@ref(fig:methodsfig)). In the dichoptic condition the non-target eye saw a fixed temporal contrast of 48\%. In the remaining three conditions (termed the cross-frequency conditions) one eye's disc flickered at 1.6Hz, and the other eye's disc flickered at 2Hz. We also tested monocular (one eye sees 1.6Hz flicker, the other sees mean luminance), binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 2Hz, mask contrast of 48\% at 1.6Hz in the other eye) arrangements (see lower rows of Figure \@ref(fig:methodsfig)). The rationale for flickering both eyes at 2Hz is that we can then measure summation behaviour between the eyes in the pupil and EEG response at 2Hz. The rationale for flickering the eyes at different frequencies is that this permits measurement of suppression between the eyes (i.e. the reduction in the 2Hz response when a 1.6Hz mask component is added to the other eye). We counterbalanced presentation of the target stimulus across the left and right eyes.

```{r methodsfig, fig.cap="Schematic diagram illustrating the ocular arrangements, and temporal waveforms of the luminance modulations used in Experiment 1. Shaded waveforms indicate a target stimulus, that was presented at one of five contrasts on each trial (denoted by the shading levels). Unshaded waveforms indicate mask stimuli, which were presented at a fixed contrast level of 48\\% regardless of the target contrast. Each waveform corresponds to a 1 second period of a 12 second trial, and coloured symbols are for consistency with Figures 2 and 3. The icon in the upper left corner illustrates the stimulus appearance (a luminous disc against a black background). The left and right eye assignments were counterbalanced across trials in the experiment (i.e. the monocular stimulus could be shown to either eye with equal probability).", fig.align="center", echo=FALSE}

knitr::include_graphics(paste0(figdir,'methodsfig.pdf'))

@@

```{r createpupilfig, include=FALSE, results='hide'}

@@

```{r pupildata, fig.cap="Summary of pupillometry results for N=30 participants. Panel (a) shows a group average waveform for binocular presentation (low pass filtered at 5Hz), with the driving signal plotted at the foot. Negative values indicate constriction relative to baseline, and positive values indicate dilation. Panel (b) shows the average Fourier spectrum (absolute amplitude values). Panels (c,d) show contrast response functions for pupil diameter at 2Hz for different conditions. Panel (e) shows contrast response functions at 1.6Hz for three conditions. Shaded regions and error bars indicate bootstrapped standard errors.", fig.align="center", echo=FALSE}

@@

```{r createEEGfig, include=FALSE, results='hide'}

@@

```{r createBSfig, include=FALSE, results='hide'}

@@

```{r createTFfig, include=FALSE, results='hide'}

@@

The results from the matching experiment are shown in Figure \@ref(fig:matchingdata). Each data point indicates the contrast levels required in each eye that were perceptually equivalent to the binocular 24\% (red circles) and 48\% (blue circles) matching contrasts. At both matching contrasts, we see a very substantial increase in the physical contrast required for a monocular target (data points along the *x*- and *y*-axes), compared to a binocular target (points along the diagonal of *x*=*y*). For example with a 48\% match, the monocular targets required contrasts close to 100\%, whereas binocular targets required a contrast of around 50\%. The data points between these extremes also fall close to the predictions of a linear summation model (diagonal dotted lines), and are inconsistent with a winner-takes-all (or MAX) model (dashed lines). Overall, these matching results are consistent with the approximately linear summation effects observed in the EEG data of Experiment 1 (Figure \@ref(fig:EEGdata)c,f).

### 2023-06-20 manuscript.tex
Added new figure

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink (\protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}). There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; Wyatt and Musselman (\protect\hyperlink{ref-Wyatt1981}{1981})). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus (\protect\hyperlink{ref-Angee2021}{Angée et al., 2021}; \protect\hyperlink{ref-Mathot2018}{Mathôt, 2018}; \protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}; \protect\hyperlink{ref-Wang2015}{Wang and Munoz, 2015}). To account for the consensual response, these brain regions must combine information from the left and right eyes (\protect\hyperlink{ref-Doesschate1967}{Doesschate and Alpern, 1967}), yet the computation that achieves this is achieved is unclear.

@@

The stimuli were two discs of achromatic flickering light with a diameter of 3.74 degrees, presented on a black background. The same stimuli were used for all three experiments. Four dark red lines were added around both discs to help with their perceptual fusion, giving the appearance of a single binocular disc (see insert in Figure \ref{fig:pupildata}b for an example of the fused stimulus). The discs were viewed through a four-mirror stereoscope, which used front silvered mirrors to avoid internal reflections, and meant that participants saw a single fused disc. The use of a stereoscope allowed us to modulate the stimuli in three different ocular configurations: monocular, binocular, and dichoptic. Note that during monocular presentation of flicker, the unstimulated eye still saw the static (non-flickering) disc of mean luminance.

@@

We included six distinct ocular conditions, each at five temporal contrast levels (combined factorially) relative to the mean luminance: 6, 12, 24, 48 and 96\%. Contrast was defined as temporal Michelson contrast; the difference between maximum and minimum luminances, scaled by the mean and expressed as a percentage. In the first three conditions, the discs flickered at 2 Hz, in either a monocular, binocular, or dichoptic arrangement. In the dichoptic condition the non-target eye saw a fixed temporal contrast of 48\%. In the remaining three conditions (termed the cross-frequency conditions) one eye's disc flickered at 1.6Hz, and the other eye's disc flickered at 2Hz. We also tested monocular (one eye sees 1.6Hz flicker, the other sees mean luminance), binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 2Hz, mask contrast of 48\% at 1.6Hz in the other eye) arrangements. The rationale for flickering both eyes at 2Hz is that we can then measure summation behaviour between the eyes in the pupil and EEG response at 2Hz. The rationale for flickering the eyes at different frequencies is that this permits measurement of suppression between the eyes (i.e.~the reduction in the 2Hz response when a 1.6Hz mask component is added to the other eye). We counterbalanced presentation of the target stimulus across the left and right eyes.

@@

\caption{Summary of pupillometry results for N=30 participants. Panel (a) shows a group average waveform for binocular presentation (low pass filtered at 5Hz), with the driving signal plotted at the foot. Negative values indicate constriction relative to baseline, and positive values indicate dilation. Panel (b) shows the average Fourier spectrum (absolute amplitude values), with an inset image illustrating the stimulus appearance (upper right). Panels (c,d) show contrast response functions for pupil diameter at 2Hz for different conditions. Panel (e) shows contrast response functions at 1.6Hz for three conditions. Shaded regions and error bars indicate bootstrapped standard errors.}\label{fig:pupildata}
\end{figure}

@@

The results from the matching experiment are shown in Figure \ref{fig:matchingdata}. Each data point indicates the contrast levels required in each eye that were perceptually equivalent to the binocular 24\% (red circles) and 48\% (blue circles) matching contrasts. At both matching contrasts, we see a very substantial increase in the physical contrast required for a monocular target (data points along the x- and y-axes), compared to a binocular target (points along the diagonal of x=y). For example with a 48\% match, the monocular targets required contrasts close to 100\%, whereas binocular targets required a contrast of around 50\%. The data points between these extremes also fall close to the predictions of a linear summation model (diagonal dotted lines), and are inconsistent with a winner-takes-all (or MAX) model (dashed lines). Overall, these matching results are consistent with the approximately linear summation effects observed in the EEG data of Experiment 1 (Figure \ref{fig:EEGdata}c,f).

>>>>

The pupillary light reflex is an automatic constriction of the iris sphincter muscles in response to increases in light levels, which causes the pupil to shrink (\protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}). There is a clear binocular component to this reflex, as stimulation of one eye still causes constriction of the other eye's pupil (termed the consensual response; Wyatt and Musselman (\protect\hyperlink{ref-Wyatt1981}{1981})). Importantly, the neuroanatomical pathway involved completely bypasses the canonical cortical pathway (retina to V1), instead involving a network of subcortical nuclei, including the Pretectal Olivary nucleus, Superior Cervical ganglion, and Edinger-Westphal nucleus (\protect\hyperlink{ref-Angee2021}{Angée et al., 2021}; \protect\hyperlink{ref-Mathot2018}{Mathôt, 2018}; \protect\hyperlink{ref-McDougal2008}{McDougal and Gamlin, 2008}; \protect\hyperlink{ref-Wang2015}{Wang and Munoz, 2015}). To account for the consensual response, these brain regions must combine information from the left and right eyes (\protect\hyperlink{ref-Doesschate1967}{Doesschate and Alpern, 1967}), yet the computation that achieves this is achieved is unclear. The pupil response is temporally low-pass (\protect\hyperlink{ref-Spitschan2014}{Spitschan et al., 2014}), most likely due to the mechanical limitations of the iris sphincter and dilator muscles.

@@

The stimuli were two discs of achromatic flickering light with a diameter of 3.74 degrees, presented on a black background. The same stimuli were used for all three experiments. Four dark red lines were added around both discs to help with their perceptual fusion, giving the appearance of a single binocular disc (see upper left insert in Figure \ref{fig:methodsfig} for an example of the fused stimulus). The discs were viewed through a four-mirror stereoscope, which used front silvered mirrors to avoid internal reflections, and meant that participants saw a single fused disc. The use of a stereoscope allowed us to modulate the stimuli in three different ocular configurations: monocular, binocular, and dichoptic. Note that during monocular presentation of flicker, the unstimulated eye still saw the static (non-flickering) disc of mean luminance.

@@

We included six distinct ocular conditions, each at five temporal contrast levels (combined factorially) relative to the mean luminance: 6, 12, 24, 48 and 96\%. Contrast was defined as temporal Michelson contrast; the difference between maximum and minimum luminances, scaled by the mean and expressed as a percentage. In the first three conditions, the discs flickered at 2 Hz, in either a monocular, binocular, or dichoptic arrangement (see upper rows of Figure \ref{fig:methodsfig}). In the dichoptic condition the non-target eye saw a fixed temporal contrast of 48\%. In the remaining three conditions (termed the cross-frequency conditions) one eye's disc flickered at 1.6Hz, and the other eye's disc flickered at 2Hz. We also tested monocular (one eye sees 1.6Hz flicker, the other sees mean luminance), binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 2Hz, mask contrast of 48\% at 1.6Hz in the other eye) arrangements (see lower rows of Figure \ref{fig:methodsfig}). The rationale for flickering both eyes at 2Hz is that we can then measure summation behaviour between the eyes in the pupil and EEG response at 2Hz. The rationale for flickering the eyes at different frequencies is that this permits measurement of suppression between the eyes (i.e.~the reduction in the 2Hz response when a 1.6Hz mask component is added to the other eye). We counterbalanced presentation of the target stimulus across the left and right eyes.

{\centering \includegraphics{Figures/methodsfig}

\caption{Schematic diagram illustrating the ocular arrangements, and temporal waveforms of the luminance modulations used in Experiment 1. Shaded waveforms indicate a target stimulus, that was presented at one of five contrasts on each trial (denoted by the shading levels). Unshaded waveforms indicate mask stimuli, which were presented at a fixed contrast level of 48\% regardless of the target contrast. Each waveform corresponds to a 1 second period of a 12 second trial, and coloured symbols are for consistency with Figures 2 and 3. The icon in the upper left corner illustrates the stimulus appearance (a luminous disc against a black background). The left and right eye assignments were counterbalanced across trials in the experiment (i.e. the monocular stimulus could be shown to either eye with equal probability).}\label{fig:methodsfig}
\end{figure}

@@

\caption{Summary of pupillometry results for N=30 participants. Panel (a) shows a group average waveform for binocular presentation (low pass filtered at 5Hz), with the driving signal plotted at the foot. Negative values indicate constriction relative to baseline, and positive values indicate dilation. Panel (b) shows the average Fourier spectrum (absolute amplitude values). Panels (c,d) show contrast response functions for pupil diameter at 2Hz for different conditions. Panel (e) shows contrast response functions at 1.6Hz for three conditions. Shaded regions and error bars indicate bootstrapped standard errors.}\label{fig:pupildata}
\end{figure}

@@

The results from the matching experiment are shown in Figure \ref{fig:matchingdata}. Each data point indicates the contrast levels required in each eye that were perceptually equivalent to the binocular 24\% (red circles) and 48\% (blue circles) matching contrasts. At both matching contrasts, we see a very substantial increase in the physical contrast required for a monocular target (data points along the \emph{x}- and \emph{y}-axes), compared to a binocular target (points along the diagonal of \emph{x}=\emph{y}). For example with a 48\% match, the monocular targets required contrasts close to 100\%, whereas binocular targets required a contrast of around 50\%. The data points between these extremes also fall close to the predictions of a linear summation model (diagonal dotted lines), and are inconsistent with a winner-takes-all (or MAX) model (dashed lines). Overall, these matching results are consistent with the approximately linear summation effects observed in the EEG data of Experiment 1 (Figure \ref{fig:EEGdata}c,f).
